Structural Insights 2023: A Linear Analysis of Apartment Age and Safety Scores in Toronto

### 2024-03-14 paper/paper.qmd
updated data and model

```{r setup, include=FALSE, message=FALSE}
## Import necessary libraries
library(ggplot2)
library(dplyr)
library(tidyverse)
library(here)
library(readr)
```

## Introduction

@@

## Data

@@

@@

### Variable Description

@@

### Data Analysis Tools

### Data Exportation

After cleaning, the modified dataset, now stored in the `cleaned_data` data frame, is written to a new CSV file named 'cleaned_building_data.csv'. This file is saved in the 'outputs/data' directory, which is again constructed using the `here` function for path management. Saving the cleaned dataset allows for a stable version of the dataset to be used for analysis or sharing, ensuring reproducibility of results.

Statistically, this script is preparing the dataset for analysis by ensuring the integrity and relevance of the data, which is foundational for any reliable statistical analysis. By creating the 'AGE' column, the script enables researchers to analyze the relationship between the age of a building and other variables of interest, such as compliance with property standards, in a quantitative manner.

## Model

@@

### Model set-up

@@

\begin{align}
y_i &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times \mbox{AGE}_i \\
\beta_0 &\sim \mbox{Normal}(90, 10) \\
\beta_1 &\sim \mbox{Normal}(-0.1, 0.05)
\end{align}

The model is executed in R using the lm function. We select normal priors for the intercept and slope based on initial exploratory data analysis.

### Model justification

A negative coefficient for AGE is anticipated, hypothesizing that older buildings may have lower safety scores due to factors such as wear and deterioration over time. We anticipate the intercept to represent the evaluation score for a newly built property, with the slope capturing the average yearly decrease in score.

## Result

```{r echo=FALSE}
# Retrieve the linear model from the saved .rds file
lm_model <- readRDS(file = here("models", "linear_model.rds"))
```

### Model Coefficients Interpretation

```{r echo=FALSE}
# Extract coefficients from the model
coefficients <- summary(lm_model)$coefficients
coefficients
```
The regression model provides two key coefficients to understand the relationship between building age and the evaluation score:

**Intercept** ($\beta_0$)  
The model estimates the intercept, $\beta_0$, to be 93.61663. This is the predicted evaluation score for a building at the time it is newly constructed, where AGE equals zero. Accompanied by a standard error of 0.64805, this coefficient's estimate signals the level of precision with which we can predict the evaluation score for new buildings. The t-value associated with the intercept is 144.459, which is highly significant, as indicated by the p-value of practically zero. This signifies a very strong statistical assurance in the intercept's estimate.

**Slope** ($\beta_1$)  
The slope of the model, $\beta_1$, is determined to be -0.10079. This value quantifies the expected change in the building evaluation score for each additional year of age; specifically, the score is projected to decrease by about 0.10079 points annually. The standard error of 0.01015 suggests high precision in this estimation. The t-value of -9.929 reaffirms the reliability of this coefficient, with a corresponding p-value of 1.226807e-22, far below the 0.05 threshold, confirming the impact of age on evaluation scores is statistically significant.

These coefficients paint a picture of gradual decline in building quality as measured by the evaluation scores over time. The statistical significance of both coefficients assures us that the model's predictions are based on real effects and not on random fluctuations within the dataset. This analysis emphasizes the need for rigorous maintenance and periodic updates to older buildings to preserve their integrity and safety.

### Model Equation

\begin{align}
y_i &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times \mbox{AGE}_i \\
\beta_0 &= 93.6166 \\
\beta_1 &= -0.1008
\end{align}

In this model, $y_i$ represents the predicted evaluation score for the $i$-th building, $\mu_i$ is the mean of the normal distribution for the $i$-th observation, $\beta_0$ is the intercept, and $\beta_1$ is the slope representing the change in evaluation score per year of building age.

### Analysis of Residuals

```{r echo=FALSE}
# Extract residuals from the model
residuals <- resid(lm_model)
summary(residuals)
```

**1. Minimum Residual**: The minimum residual is -69.662, indicating that there is at least one building for which the model's predicted evaluation score was lower than the actual score by approximately 69.662 points. This could be an outlier or a building with exceptional characteristics not captured by the model.

**2. First Quartile Residual**: The first quartile (Q1) of the residuals is -4.364, meaning that 25% of the buildings have evaluation scores that are less than 4.364 points below the model's prediction for their age.

**3. Median Residual**: The median residual is 1.702, suggesting a slight positive bias in the model's predictions, as the median of the error distribution is above zero. This indicates that, on average, the model tends to slightly overestimate the evaluation scores.

**4. Third Quartile Residual**: The third quartile (Q3) is 5.833, indicating that 75% of the buildings have evaluation scores within 5.833 points of the model's predictions or better. This shows that the model provides reasonably accurate predictions for the majority of buildings.

**5. Maximum Residual**: The maximum residual is 15.859, suggesting that there is at least one building for which the model's predicted evaluation score was higher than the actual score by approximately 15.859 points. This could be another outlier or a building that performed exceptionally well.

@@

### Plot for the linear model
```{r fig-linear-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="plot for the linear relationship between building age and evaluation score"}
# Create a scatter plot with a regression line to visualize the relationship
improved_plot <- ggplot(cleaned_data, aes(x = AGE, y = CURRENT.BUILDING.EVAL.SCORE)) +
  geom_point(alpha = 0.7, size = 1) +  # Adjust transparency and size of points
  geom_smooth(method = "lm", color = "blue", size = 1) +  # Change the size of the regression line
  labs(
    x = "Building Age",
    y = "Current Building Evaluation Score",
  ) +
  theme_minimal() +  # Use a minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8), # Rotate x axis text and adjust horizontal justification
    axis.text.y = element_text(size = 8),
    legend.position = "none", # Hide the legend if not necessary
    plot.margin = margin(1, 1, 1, 1, "cm") # Adjust the margin around the plot
  ) +
  theme(legend.position = "bottom") # Move the legend to the bottom to prevent it from overlapping the plot

# Print the improved plot
print(improved_plot)
```

@@

## Discussion

### Overview of the Study

@@

### Insights into Urban Development and Maintenance

@@

### Policy Implications and Housing Quality

@@

### Limitations of the Study

@@

### Future Directions

@@

## Appendix

## References

>>>>

```{r setup, include=FALSE, message=FALSE}
## Import necessary libraries
library(ggplot2)
library(dplyr)
library(tidyverse)
library(here)
library(readr)
library(knitr)
library(kableExtra)
```

# Introduction

@@

# Data

@@

## Raw Data

@@

## Data Analysis Tools

## Variable Description

@@

## Sample of Cleaned Building Evaluation Data

```{r}
#| label: tbl-data-sample
#| tbl-cap: Sample of Building Evaluation Data
#| echo: false
#| 
# Select the first few rows of the dataset and the specific columns
sample_data <- head(cleaned_data) %>%
  select(X_id, AGE, CURRENT.BUILDING.EVAL.SCORE)

# Create a nicely formatted table
kable(sample_data)
```
@tbl-data-sample represents a subset of the larger building evaluation data. Each row corresponds to a unique building identified by X_id. The AGE column indicates the age of the building in years, while the CURRENT.BUILDING.EVAL.SCORE column shows the building's current evaluation score on a scale from 0 to 100, with higher scores indicating better compliance with safety standards.

## Data Exploration:

### Data Summary
```{r}
#| label: tbl-summary-table
#| tbl-cap: Summary Statistics for the Cleaned Dataset
#| echo: false

summary_data <- cleaned_data %>%
  summarise(
    Count = n(),
    Mean_Age = mean(AGE, na.rm = TRUE),
    Std_Age = sd(AGE, na.rm = TRUE),
    Mean_Score = mean(CURRENT.BUILDING.EVAL.SCORE, na.rm = TRUE),
    Std_Score = sd(CURRENT.BUILDING.EVAL.SCORE, na.rm = TRUE)
  ) %>%
  mutate(across(everything(), ~round(., 3)))

# Create and print a nicely formatted table
summary_table <- kable(summary_data, 
                       col.names = c("Count", "Mean Age", "SD of Age", 
                                     "Mean Score", "SD of Score"), 
                       align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "center") %>%
  column_spec(1, bold = TRUE) # Make the first column bold

# Print the summary table
summary_table
```

@tbl-summary-table presents a concise overview of the key descriptive metrics for the age of buildings and their evaluation scores within the dataset:

- **Count**: The dataset encompasses a total of 1,748 buildings, indicating a robust sample size for statistical analysis.

- **Mean Age**: The average age of the buildings is approximately 60.6 years, suggesting that the dataset primarily includes buildings that have been standing for over half a century.

- **Standard Deviation of Age**: With a standard deviation of about 19.9 years, there is considerable variability in the ages of the buildings. This range implies a diverse sample, encompassing both relatively new constructions and much older structures.

- **Mean Score**: The average building evaluation score is approximately 87.5 on a scale from 0 to 100. A higher score corresponds to better adherence to safety and maintenance standards, suggesting that the buildings in this dataset are generally well-maintained.

- **Standard Deviation of Score**: The standard deviation of approximately 8.7 for the evaluation scores indicates that while scores are clustered around a high mean, there is still notable variability, reflecting differences in the condition or compliance of the buildings.

These statistics offer a snapshot of the dataset's characteristics, providing foundational insights for further analysis regarding the state of building safety and integrity in the population studied.

### Age Distribution
```{r fig-age-distribution, echo=FALSE, fig.cap="Distribution of building's age for the cleaned data set"}
# Histogram for the distribution of building age
ggplot(cleaned_data, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    x = "Age of Building (Years)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

@fig-age-distribution illustrates the frequency distribution of the ages of buildings within the dataset. The x-axis represents the age of the buildings in years, while the y-axis shows the frequency of buildings within each age range. The distribution appears to be unimodal with a peak around the 50-year mark, indicating that a substantial number of the buildings within this dataset were constructed around half a century ago. Following this peak, the frequency gradually decreases, showing that fewer buildings have been in existence beyond this age. The distribution suggests a relatively young building population, with ages concentrated below 100 years, and very few buildings reaching beyond 120 years of age. This visualization aids in understanding the age profile of buildings under consideration for safety evaluation scoring.

### Score Distribution

```{r fig-score-distribution, echo=FALSE, fig.cap="Distribution of building's evaluation score for the cleaned data set"}
# Histogram for CURRENT.BUILDING.EVAL.SCORE distribution
ggplot(cleaned_data, aes(x = CURRENT.BUILDING.EVAL.SCORE)) +
  geom_histogram(binwidth = 5, fill = "darkgreen", color = "white") +
  labs(
    x = "Current Building Evaluation Score",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

@fig-score-distribution showcases the distribution of building evaluation scores. The x-axis represents the scores assigned to buildings, which can range from 0 to 100, with the y-axis displaying the frequency of buildings for each score interval. The chart reveals a concentration of high evaluation scores, suggesting that many of the buildings in the dataset are rated as being in good or excellent condition according to the evaluation criteria. The tallest bars are clustered towards the higher end of the score range, indicating that buildings with scores close to 100 are prevalent. This positive skew in the distribution might reflect effective building standards and maintenance practices in place. The spread of the scores and their frequency can offer insights into the overall health of the evaluated buildings, influencing urban planning and policy development.

# Model

@@

## Model set-up

@@

\begin{align}
y_i &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times x_i \\
\beta_0 &\sim \mbox{Normal}(90, 10) \\
\beta_1 &\sim \mbox{Normal}(-0.1, 0.05)
\end{align}

The model is implemented in R using the rstanarm package, which allows us to incorporate prior beliefs about the parameters and to estimate the posterior distributions. For the intercept $\beta_0$ and slope $\beta_1$ , we choose normal prior distributions centered around our initial estimates based on exploratory data analysis. The `stan_glm` function from `rstanarm` is utilized to carry out the Bayesian regression, providing a framework for inference that incorporates both the data and our prior knowledge.

The priors are chosen to reflect reasonable beliefs about the parameters before observing the data. The intercept prior reflects an expectation that a new building will have a high safety score around 90, with a wide spread to accommodate uncertainty. Similarly, the slope prior indicates an expected small negative impact of age on the safety score, acknowledging some uncertainty in the precise effect. The `stan_glm` function combines these priors with the data to estimate the posterior distributions of the model parameters.

## Model justification

We hypothesize that the coefficient for building age ($\beta_1$) will be negative, reflecting the tendency for older buildings to exhibit lower safety scores. This expectation is rooted in the general understanding that buildings can deteriorate over time due to wear, aging materials, and possible obsolescence in older construction methods. The intercept ($\beta_0$), representing the baseline evaluation score for a new building, is presumed to be relatively high, indicating the score a building would receive if it were built in the current period with modern standards. The slope ($\beta_1$) is therefore of particular interest, as it encapsulates the average annual decline in evaluation score attributable to aging. The Bayesian framework allows us to incorporate these hypotheses as priors in our model, thereby aligning the analysis with our substantive expectations and providing a more nuanced interpretation of the data.

# Result

```{r echo=FALSE}
# Retrieve the Bayesian linear model from the saved .rds file
bayesian_model <- readRDS(file = here("models", "bayesian_linear_model.rds"))
```

## Model Coefficients Interpretation

```{r echo=FALSE}
# Extract coefficients from the Bayesian model
coefficients <- coef(bayesian_model)
print(coefficients)
```
The regression model provides two key coefficients to understand the relationship between building age and the evaluation score:

**Intercept** ($\beta_0$):  
The model's intercept, $\beta_0$, with a mean posterior estimate of 93.55985, represents the baseline evaluation score for a hypothetical newly constructed building. In Bayesian terms, this value is not a single point estimate but rather the mean of the posterior distribution for the intercept. This distribution reflects our belief about the probable values of the evaluation score for a new building, informed by the data and our prior assumptions.

**Slope** ($\beta_1$):  
Similarly, the slope parameter $\beta_1$ has a mean posterior estimate of -0.10059. It describes the average decrease in the evaluation score associated with each additional year of a building's age. Again, this is the mean of the posterior distribution for the slope, encompassing both our prior beliefs and the observed data. A mean estimate that is negative suggests a trend where older buildings have lower evaluation scores.

**Interpretation**:
The Bayesian analysis suggests that there is a likely negative relationship between building age and evaluation score. The mean estimates of the intercept and slope, along with their posterior distributions, indicate that newer buildings tend to have higher evaluation scores, which decrease as buildings age. This trend highlights the importance of age as a factor in building evaluations and underscores the need for ongoing maintenance and updates to preserve building safety and integrity over time.

## Model Equation

\begin{align}
y_i &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times x_i \\
\beta_0 &= 93.55985 \\
\beta_1 &= -0.10059
\end{align}

In this model, $y_i$ represents the predicted evaluation score for the $i$-th building, $x_i$ represents the age for the $i$-th building, $\mu_i$ is the mean of the normal distribution for the $i$-th observation, $\beta_0$ is the intercept, and $\beta_1$ is the slope representing the change in evaluation score per year of building age.

## Analysis of Residuals

```{r echo=FALSE}
# Extract residuals from the Bayesian model
residuals <- residuals(bayesian_model)
summary(residuals)
```

**1. Minimum Residual**: The minimum residual is approximately -69.619, indicating that there is at least one building for which the model's predicted evaluation score was lower than the actual score by approximately 69.662 points. This could be an outlier or a building with exceptional characteristics not captured by the model.

**2. First Quartile Residual**: The first quartile (Q1) of the residuals is approximately -4.320, meaning that 25% of the buildings have evaluation scores that are less than 4.364 points below the model's prediction for their age.

**3. Median Residual**: The median residual is approximately 1.757, suggesting a slight positive bias in the model's predictions, as the median of the error distribution is above zero. This indicates that, on average, the model tends to slightly overestimate the evaluation scores.

**4. Third Quartile Residual**: The third quartile (Q3) is approximately 5.878, indicating that 75% of the buildings have evaluation scores within 5.833 points of the model's predictions or better. This shows that the model provides reasonably accurate predictions for the majority of buildings.

**5. Maximum Residual**: The maximum residual is approximately 15.896, suggesting that there is at least one building for which the model's predicted evaluation score was higher than the actual score by approximately 15.896 points. This could be another outlier or a building that performed exceptionally well.

@@

## Plot for the linear model
```{r fig-linear-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="plot for the linear relationship between building age and evaluation score"}
# Enhanced scatter plot with regression line
improved_plot <- ggplot(cleaned_data, aes(x = AGE, y = CURRENT.BUILDING.EVAL.SCORE)) +
  geom_point(alpha = 0.7, size = 1) +  # Adjust transparency and size of points
  geom_smooth(method = "lm", color = "blue", size = 1) +  
  labs(
    x = "Building Age",
    y = "Building Evaluation Score",
    title = "Impact of Building Age on Evaluation Score"
  ) +
  theme_minimal() +  # Use a minimal theme
  theme(
    plot.title = element_text(size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    legend.position = "none"  # Hide the legend if not necessary
  )

# Display the plot
print(improved_plot)
```

@@

# Discussion

## Overview of the Study

@@

## Insights into Urban Development and Maintenance

@@

## Policy Implications and Housing Quality

@@

## Limitations of the Study

@@

## Future Directions

@@

# Appendix

# References
